EDIpack2: interoperable Lanczos-based solver for generic quantum impurity problems

### 2025-05-02 05_examples.tex
Updated draft
- use subfiles, so we can compile in every subfile
- written something re. umatrix
- updated gitignore
Revised MIT section. No multi-band attractive with replica bath -> directly to interacting BHZ + replica for triplet exctions.

\subsection{Bethe lattice DMFT}\label{SecExamplesBetheDMFT}
The description of the Mott or metal to insulator transition (MIT)
within the Bethe lattice, i.e. Cayley tree with connection
$z\to\infty$ is conventionally considered the test bed of any DMFT
application. Here  we discuss a guided implementation based on the
Fortran API of \NAME of the DMFT solution of the Bethe lattice at
$T=0$. 
We consider a Fermi-Hubbard model:
$$
H = -t \sum_{\langle ij\rangle\s}c^+_{i\s}c_{j\s} + U \sum_i n_{i\up}n_{i\dw}
$$
where $c^+_{i\s}$ ($c_{i\s}$) is the creation (destruction) operator for an electron at site $i$
and with spin $\s$, $n_{i\s}=c^+_{i\s}c_{i\s}$ is the occupation
operator and the first sum is over the nearest neighbor sites $\langle
ij\rangle$. 
The system is defined on a Bethe lattice with density of states
$\rho(x)=\frac{1}{2D}\sqrt{D^2-x^2}$, where $D=2t$ is the
half-bandwidth~\cite{Georges1996}.
In DMFT this model is mapped
onto a quantum impurity problem with an effective electronic bath
which needs to be determined self-consistently.
\begin{lstlisting}[style=fstyle,numbers=none,basicstyle={\scriptsize\ttfamily}]
program ed_hm_bethe
   USE EDIPACK2
   USE SCIFOR
   implicit none
   !> Number of discretization points for the Bethe DOS
   integer,parameter                           :: Le=5000
   !> Bethe half-bandwidth = energy unit
   real(8),parameter                           :: D=1d0
   !>Bath:
   real(8),allocatable                         :: Bath(:)
   !> local dynamical functions, rank-5 [Nspin,Nspin,Norb,Norb,L]
   complex(8),allocatable,dimension(:,:,:,:,:) :: Weiss,Smats,Sreal,Gmats,Greal,Weiss_
   ... (omissis)
   !> READ THE input using EDIpack procedure:
   call ed_read_input(trim(finput))
   !
   !> Setup the Bethe DOS $\rho$ (linear dispersion)
   Ene = linspace(-D,D,Le,mesh=de)
   DOS = dens_bethe(Ene,D)
\end{lstlisting}

@@

\paragraph{Results}
In the following we present some results obtained by executing this
simple program varying the interaction strenght $U$. 
Differently from the previous case of a quantum impurity embedded in a
given bath describing the progressive formation of a
strongly renormalized Fermi liquid state, here the DMFT
self-consistency allows to describe the transition from a correlated
metal to a Mott insulating state.

To illustrate this point, in panel **A** we report the evolution of
the spectral function $-{\rm Im}G(\omega)/\pi$ as a function of
$U$. Despite the *spiky* nature of the spectrum, due to the
finite size (i.e. number of poles) of the discretized effective bath,
one can clearly distinguish the renormalization of the central
quasi-particle peak at low-energy and the concomitant formation of
rather incoherent  high-energy features which will develop into
Hubbard bands for $U>U_c$, with $U_c\simeq 2.8D$.

@@

In the panels (B) and (C) we further discuss the metal-insulator
transition by showing the evolution of the self-energy functions.
In panel (B) we report the self-energy ${\rm Im}\Sigma(i\omega)$
on the Matsubara axis in the low energy regime. Increasing $U$
we observe the progressive growth of this function until it takes a
diverging behavior crossing the critical interaction strenght. We
recall that this behavior can be observed on the Matsubara axis
because of the particle-hole symmetry of the problem.

Using the relation:

$$
   \frac{\Im\Sigma(i\omega_n)}{\omega_n}_{|_{\omega_n\rightarrow 0}}=
   \frac{1}{\pi}\int_{\mathbb R}d\epsilon \frac{\Re\Sigma(\epsilon)}{\epsilon^2}=
   \frac{\partial\Re\Sigma}{\partial\omega}_{|_{\omega\rightarrow 0}}.
$$
we can extract the quasi-particle renormalization constant :math:`Z`
from the linear behavior of ${\rm Im}\Sigma(i\omega)$ near
$\omega=0$ in the metallic regime. The results are highlighted
in the figure and the values of $Z$ are reported in the legend.

Using the right hand side of the previous relation, we show in
panel (C) the behavior of  ${\rm Re}\Sigma(\omega)$
on the real axis around the Fermi level. Again, increasing $U$
we observe the slope of the linear behavior to increase until the
critical point is crossed and an insulating state is reached. On the
real-axis this is signaled by the divergence of the imaginary part of
the self-energy   ${\rm Im}\Sigma(\omega) \rightarrow -\infty$
near the chemical potential, which here is set to zero by
particle-hole symmetry. The corresponding real part shows a
discontinuity visibile in the panel (C).

@@

The iterative scheme for the solution of DMFT closely follows the
sequence discussed in \secu{SecExamplesBetheDMFT}:   
\begin{enumerate}
\item Call the exact diagonalization {\bf impurity solver} {\tt
    ed.solve} providing the set of bath parameters as input. 
\item Using the dedicated
  input/output procedures {\tt ed.get\_sigma} we retrieve the self-energy functions  
  $\hat{\Sigma}(i\omega)$ and $\hat{S}(i\omega)$ on the 
  Matsubara axis.
\item
  Evaluate the interacting local Green's functions $\hat{G}_{loc}$ and
  $\hat{F}_{loc}$:
  \begin{equation}
  {\mathbf G}_{loc}(i\omega) =
  \int_\RRR d\e \rho(\e)
  \begin{pmatrix}
    (i\omega +\mu)\hat{\11} - \hat{h}^0 - \hat{\Sigma}(i\omega) -\e & -\hat{S}(i\omega) \\
    -\hat{S}(i\omega) & (i\omega +\mu)\hat{\11} + \hat{h}^0 +
    \hat{\Sigma}^*(i\omega) +\e\\
  \end{pmatrix}^{-1}
\end{equation}
\item Update the normal and anomalous Weiss fields, respectively 
  $\GG_0^{-1}$ and $\FF_0^{-1}$, through the {\bf self-con\-sis\-ten\-cy}
    relation: $\mathbfcal{G}^{-1}_0(i\omega) = {\mathbf G}_{loc}(i\omega) +
    {\mathbf \Sigma}(i\omega)$ in Nambu space. 
  \item Optimize the bath parameters  to best describe the  updated
    Weiss fields using \NAME provided conjugate gradient  fit
    procedure {\tt ed.chi2\_fitgf}.
\end{enumerate}
which is implemented in Python as follows:
\begin{lstlisting}[style=mypython,numbers=none,basicstyle={\scriptsize\ttfamily}]
#DMFT CYCLE
converged=False;iloop=0
while (not converged and iloop<ed.Nloop ):
iloop=iloop+1

@@

To beging with, we report in  panel (A) of \figu{figEx2} the
evolution of the spectral density, obtained from the local normal
Green's function $-\tfrac{1}{\pi}\Im{G}_{loc}(\omega)$, as a function
of the attraction $U$. For any finite value of  $U$ the Van Hove peak
at $\omega=0$ of the 2D square lattice (see $U=0$
solution) gets split by the formation of a superconductive gap of
width which increases with $U$.
The existence of a gap is associated to the formation of a finite
order parameter $\phi=\langle c_\up c_\dw\rangle$. The behavior of
this quantity, shown in panel (B), highlights crossover from the BCS
regime at weak coupling to a BEC one for large $U$.
The former regime is characterized by the well known exponential increase of
$\phi$, which is computationally the hardest regime to adress. At
strong coupling or BEC regime the order parameter saturates to its
limiting value $\phi\to 1/2$. The comparison with mean-field result
(gray) underlines the effect of local quantum fluctuations in slightly
suppressing the order parameter. This effect is particularly active in
the intermediate regime, as captured by the behavior of the
correlation strength
$\theta=\frac{|S(i\omega\to9)-S(i\omega\to\infty)|}{S(i\omega\to\infty)}$
reported in panel (C). Assuming a monotonic behavior of the anomalous
self-energy function, this quantity estimates the ``strength'' of
quantum fluctuations comparing the mean-field-like behavior at high
energy scale with that near the Fermi level. A large value of $\theta$
indicates a more correlated system.
Our results show that both weak and strong coupling regime are
relatively weakly correlated, i.e. very near a mean-field solution. On
the contrary, the inter-mediate regime is characterized by a large
correlation, making the solution very different from that obtained in
BCS description.  
To further corroborate the distance between the DMFT and mean-field
solution we report in panel (D) the evolution of the double occupancy
$d=\langle n_\up n_\dw\rangle$. At half-filling the mean-field value is
determined solely by the order parameter as $d=1/4 + \phi^2$.
Our results show that the inclusion of quantum fluctuations in DMFT
slightly increase the value of $d$ in the weak coupling
regime. However, increasing the attraction the double occupations
quickly get reduced below the mean-field value up to the BEC regime.

>>>>

\subsection{Bethe lattice DMFT}\label{SecExamplesBetheDMFT}
The description of the Mott or metal to insulator transition (MIT)
within the Bethe lattice, i.e. Cayley tree with connection
$z\to\infty$ is conventionally considered the test bed of any DMFT
application. Here  we discuss a guided implementation based on the
Fortran API of \NAME of the DMFT solution of the Bethe lattice at
$T=0$. 
We consider a Fermi-Hubbard model:
$$
H = -t \sum_{\langle ij\rangle\s}c^+_{i\s}c_{j\s} + U \sum_i n_{i\up}n_{i\dw}
$$
where $c^+_{i\s}$ ($c_{i\s}$) is the creation (destruction) operator for an electron at site $i$
and with spin $\s$, $n_{i\s}=c^+_{i\s}c_{i\s}$ is the occupation
operator and the first sum is over the nearest neighbor sites $\langle
ij\rangle$. 
The system is defined on a Bethe lattice with density of states
$\rho(\e)=\frac{1}{2D}\sqrt{D^2-\e^2}$, where $D=2t$ is the
half-bandwidth~\cite{Georges1996}.
In DMFT this model is mapped
onto a quantum impurity problem with an effective electronic bath
which needs to be determined self-consistently.
\begin{lstlisting}[style=fstyle,numbers=none,basicstyle={\scriptsize\ttfamily}]
program ed_hm_bethe
   USE EDIPACK2
   USE SCIFOR
   implicit none
   !> Number of discretization points for the Bethe DOS
   integer,parameter                           :: Le=5000
   !> Bethe half-bandwidth = energy unit
   real(8),parameter                           :: D=1d0
   !>Bath:
   real(8),allocatable                         :: Bath(:)
   !> local dynamical functions, rank-5 [Nspin,Nspin,Norb,Norb,L]
   complex(8),allocatable,dimension(:,:,:,:,:) :: Weiss,Smats,Sreal,Gmats,Greal,Weiss_
   ... (omissis)
   !> READ THE input using EDIpack procedure:
   call ed_read_input(trim(finput))
   !
   !> Setup the Bethe DOS $\rho$ (linear dispersion)
   Ene = linspace(-D,D,Le,mesh=de)
   DOS = dens_bethe(Ene,D)
\end{lstlisting}

@@

\paragraph{Results}
In the following we present some results obtained for the model
introduced above. Specifically we focus on the testbed case of the interaction
driven Mott transition, descriving the progressive
transformation of a partially-filled band metallic state into a correlated insulator.

To illustrate this point, we report in panel (A) the evolution of
the spectral function $-{\rm Im}G_{loc}(\omega)/\pi$ as a function of
the interaction strenght $U$. Despite the {\it spiky} nature of the spectrum, due to the
finite size (i.e. number of poles) of the discretized effective bath,
it is possible to clearly distinguish the characteristic features of
the Mott transition. At low energy, we observe the renormalization of the 
quasi-particle peak at the Fermi level ($\omega=0$). Concomitantly the
system shows the formation of incoherent  high-energy features which
will then develop into Hubbard bands in the Mott insulating state for $U>U_c$, with $U_c\simeq 2.8D$.

@@

The formation of a spectral gap separating the Hubbard bands in the
Mott insulator is supported by the divergence, at the Fermi level, of the imaginary part of
the self-energy function. Causality imposes that the real-part of
this quantity becomes large near the singularity making it impossible
to satisfy the quasi-particles poles equations
$\omega+\mu-h^0-\e-\Re{\Sigma}(\omega)=0$, i.e. preventing coherent
quasi-particles excitations near the Fermi level. 
In the panels (B) and (C) we discuss this point  showing the evolution
of the self-energy functions $\Sigma$.
Specifically, in panel (B) we report the self-energy ${\rm Im}\Sigma(i\omega)$
on the Matsubara axis in the low energy regime. Increasing $U$
we observe the progressive growth of this function until a
diverging behavior is acheived crossing the critical interaction
strenght $U>U_c$.
We recall that this behavior can be observed on the Matsubara axis
because of the particle-hole symmetry of the problem which pins the
$\Im{\Sigma}$ singularity at $\omega=0$.
In panel (C), we report the corresponding behavior of the
$\Re{\Sigma}(\omega)$ in small energy regime around the Fermi
level. Augmenting the interaction strength $U$ leads to a sudden
increase of this self-energy component until a discontinuous behavior
is reached, associated to the divergence of the corresponding
imaginary part.

A quantitative description of the transition is obtained from the
quasi-particle renormalization constant $Z$, which correlates with the degree of
localization of the electrons being equal to 1 for a non-interacting
metal and 0 for the Mott insulator. This quantity is obtained from the
linear term in the Fermi liquid expansion of the self-energy as
$Z=(1-\tfrac{\partial\Re\Sigma}{\partial\omega}_{|_{\omega\rightarrow
    0}})^{-1}$.
Using the relation:
$$
   \frac{\Im\Sigma(i\omega_n)}{\omega_n}_{|_{\omega_n\rightarrow 0}}=
   \frac{1}{\pi}\int_{\mathbb R}d\epsilon \frac{\Re\Sigma(\epsilon)}{\epsilon^2}=
   \frac{\partial\Re\Sigma}{\partial\omega}_{|_{\omega\rightarrow 0}}.
$$
we can extract $Z$ also from the linear behavior of ${\rm Im}\Sigma(i\omega)$ for 
$\omega\to0$ in the metallic regime.

The linear fit highlighted in the panels (B)-(C) and the corresponding values of $Z$ are reported in the legend.
In both cases we observe that increasing $U$ the slope of the linear behavior of the
self-energy at low-energy increases upon increasing $U$, either on the
Matsubara or real-axis. At the transition point the slope becomes
infinite according to the divergence of the imaginary part of
the self-energy  ${\rm Im}\Sigma(\omega) \rightarrow -\infty$,
corresponding to $Z\to 0$.

@@

The iterative scheme for the solution of DMFT closely follows the
sequence discussed in \secu{SecExamplesBetheDMFT}:   
\begin{enumerate}
\item Call the exact diagonalization {\bf impurity solver} {\tt
    ed.solve} providing the set of bath parameters $\vec{x}=\{V,h\}$  as input. 
\item Using the dedicated
  input/output procedures {\tt ed.get\_sigma} we retrieve the self-energy functions  
  $\hat{\Sigma}(i\omega)$ and $\hat{S}(i\omega)$ on the 
  Matsubara axis.
\item
  Evaluate the interacting local Green's functions $\hat{G}_{loc}$ and
  $\hat{F}_{loc}$:
  \begin{equation}
  {\mathbf G}_{loc}(i\omega) =
  \int_\RRR d\e \rho(\e)
  \begin{pmatrix}
    (i\omega +\mu)\hat{\11} - \hat{h}^0 - \hat{\Sigma}(i\omega) -\e & -\hat{S}(i\omega) \\
    -\hat{S}(i\omega) & (i\omega +\mu)\hat{\11} + \hat{h}^0 +
    \hat{\Sigma}^*(i\omega) +\e\\
  \end{pmatrix}^{-1}
\end{equation}
\item Update the normal and anomalous Weiss fields, respectively 
  $\GG_0^{-1}$ and $\FF_0^{-1}$, through the {\bf self-con\-sis\-ten\-cy}
    relation: $\mathbfcal{G}^{-1}_0(i\omega) = {\mathbf G}_{loc}(i\omega) +
    {\mathbf \Sigma}(i\omega)$ in Nambu space. 
  \item Optimize the bath parameters $\vec{x}$ to best describe the  updated
    Weiss fields using \NAME provided conjugate gradient  fit
    procedure {\tt ed.chi2\_fitgf}.
\end{enumerate}
which is implemented in Python as follows:
\begin{lstlisting}[style=mypython,numbers=none,basicstyle={\scriptsize\ttfamily}]
#DMFT CYCLE
converged=False;iloop=0
while (not converged and iloop<ed.Nloop ):
iloop=iloop+1

@@

To beging with, we report in  panel (A) of \figu{figEx2} the
evolution of the spectral density, obtained from the local normal
Green's function $-\tfrac{1}{\pi}\Im{G}_{loc}(\omega)$, as a function
of the attraction $U$. For any finite value of  $U$ the Van Hove peak
at $\omega=0$ of the 2D square lattice (see $U=0$
solution) gets split by the formation of a superconductive gap of
width which increases with $U$.
The existence of a gap is associated to the formation of a finite
order parameter $\phi=\langle c_\up c_\dw\rangle$. The behavior of
this quantity, shown in panel (B), highlights crossover from the BCS
regime at weak coupling to a BEC one for large $U$.
The former regime is characterized by the well known exponential increase of
$\phi$, which is computationally the hardest regime to adress. At
strong coupling or BEC regime the order parameter saturates to its
limiting value $\phi\to 1/2$. The comparison with mean-field result
(gray) underlines the effect of local quantum fluctuations in slightly
suppressing the order parameter. This effect is particularly active in
the intermediate regime, as captured by the behavior of the
correlation strength
$\theta=\frac{|S(i\omega\to 0)-S(i\omega\to\infty)|}{S(i\omega\to\infty)}$
reported in panel (C). Assuming a monotonic behavior of the anomalous
self-energy function, this quantity estimates the ``strength'' of
quantum fluctuations comparing the mean-field-like behavior at high
energy scale with that near the Fermi level. A large value of $\theta$
indicates a more correlated system.
Our results show that both weak and strong coupling regime are
relatively weakly correlated, i.e. very near a mean-field solution. On
the contrary, the inter-mediate regime is characterized by a large
correlation, making the solution very different from that obtained in
BCS description.  
To further corroborate the distance between the DMFT and mean-field
solution we report in panel (D) the evolution of the double occupancy
$d=\langle n_\up n_\dw\rangle$. At half-filling the mean-field value is
determined solely by the order parameter as $d=1/4 + \phi^2$.
Our results show that the inclusion of quantum fluctuations in DMFT
slightly increase the value of $d$ in the weak coupling
regime. However, increasing the attraction the double occupations
quickly get reduced below the mean-field value up to the BEC regime.
